---
title: "Change scores in a synthetic training study"
author:
  - name: Workshop Participant
    affiliations:
      - name: Your University
abstract: |
  This demonstration renders a small computational analysis with a bundled
  Elsevier journal format. The data are synthetic and the result is illustrative.
keywords:
  - reproducibility
  - literate programming
  - Quarto
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
execute:
  warning: false
  message: false
---

## Introduction

Literate programming combines narrative and executable analysis [@knuth1984].
This document demonstrates that a journal format changes presentation while the
underlying calculation remains recorded in the source.

```{r}
#| include: false

library(readr)
library(dplyr)
library(ggplot2)

study <- read_csv("data/training-study.csv", show_col_types = FALSE) |>
  mutate(change = followup_score - baseline_score)
```

## Results

The synthetic dataset contains `r nrow(study)` observations.

```{r}
#| label: fig-journal-change
#| fig-cap: "Change score by training hours and study group."
#| fig-alt: "Scatterplot of training hours against change score, distinguished by study group."

ggplot(study, aes(training_hours, change, colour = group, shape = group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Training hours", y = "Change in score") +
  theme_minimal()
```

## Discussion

The output is a formatting demonstration, not evidence about a real
intervention. Journal instructions should be checked at the time of submission.

## References

