---
title: "A reproducible analysis"
subtitle: "TODO: describe the study"
author: "TODO: your name"
date: today
bibliography: bibliography.bib
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
    html-math-method: mathml
  pdf:
    number-sections: true
execute:
  warning: false
  message: false
---

# Introduction

This manuscript uses a synthetic training study to practise reproducible
writing. Literate programming keeps explanation and computation together.

<!-- TODO (Markdown activity):
- Add a Research question heading and one-sentence question.
- Add a bulleted list of aims.
- Add the displayed change-score equation.
- Cite @knuth1984.
- Add a callout noting that the data are synthetic.
-->

# Methods

The dataset contains 40 fictional participants. It is stored inside the project
so the document does not depend on an external download.

```{r}
#| label: load-and-prepare
#| include: false

# TODO (computation activity):
# 1. Load readr, dplyr, and ggplot2.
# 2. Read data/training-study.csv into an object named study.
# 3. Create change = followup_score - baseline_score.
```

<!-- TODO: Replace the typed participant count above with inline R. -->

# Results

```{r}
#| label: tbl-summary
#| tbl-cap: "Change scores by study group."
#| eval: false

# TODO:
# summary_table <- study |>
#   group_by(group) |>
#   summarise(
#     participants = n(),
#     mean_change = mean(change),
#     sd_change = sd(change)
#   )
#
# knitr::kable(summary_table, digits = 1)
```

```{r}
#| label: fig-change
#| fig-cap: "Change score by training hours and study group."
#| fig-alt: "TODO: describe the variables and visible pattern."
#| eval: false

# TODO:
# ggplot(study, aes(training_hours, change, colour = group)) +
#   geom_point(size = 2) +
#   geom_smooth(method = "lm", se = FALSE) +
#   labs(
#     x = "Training hours",
#     y = "Change in score",
#     colour = "Study group"
#   ) +
#   theme_minimal()
```

<!-- TODO: Add prose that refers to @tbl-summary and @fig-change. -->

# Discussion

TODO: Distinguish the descriptive result from a causal conclusion and name one
limitation of the synthetic study.

# Reproducibility

The source data, transformations, prose, and output instructions are stored in
one project. The completed document records session information here.

# References
