---
title: "Chunk options: runnable demonstrations"
subtitle: "Change one option, render, and compare"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: demo-setup
#| include: false

library(readr)
library(dplyr)
library(ggplot2)

study <- read_csv("data/training-study.csv", show_col_types = FALSE) |>
  mutate(change = followup_score - baseline_score)
```

# Code and output

The default shows both code and its result.

```{r}
#| label: demo-default

study |>
  summarise(participants = n(), mean_change = mean(change))
```

# `echo: false`

The code runs, but the rendered reader sees only its output.

```{r}
#| label: demo-echo
#| echo: false

study |>
  count(group)
```

# `eval: false`

The code is displayed but not executed.

```{r}
#| label: demo-eval
#| eval: false

download.file("https://example.org/large.csv", "data/large.csv")
```

# `output: false`

The code runs and creates `group_summary`, but its printed output is hidden in
this cell.

```{r}
#| label: demo-output
#| output: false

group_summary <- study |>
  group_by(group) |>
  summarise(
    participants = n(),
    mean_change = mean(change),
    sd_change = sd(change),
    .groups = "drop"
  )

group_summary
```

# A labelled table

The object created above is presented and numbered here.

```{r}
#| label: tbl-demo-summary
#| tbl-cap: "Change scores by study group."

knitr::kable(group_summary, digits = 1)
```

See @tbl-demo-summary.

# A labelled figure

```{r}
#| label: fig-demo-change
#| fig-cap: "Change score by training hours and study group."
#| fig-alt: "Scatterplot of training hours against change score. Intervention observations generally have more hours and larger changes."
#| fig-width: 7
#| fig-height: 4.5
#| dpi: 150

ggplot(study, aes(training_hours, change, colour = group, shape = group)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Training hours",
    y = "Change in score",
    colour = "Study group",
    shape = "Study group"
  ) +
  theme_minimal()
```

See @fig-demo-change.

# Inline computation

The source inserts `nrow(study)` inline rather than typing the result:

> This demonstration uses `r nrow(study)` observations and
> `r n_distinct(study$group)` groups.

# Experiments

Try one change at a time and rerender:

1. Add `echo: false` to the figure.
2. Change `fig-width` and observe the legend and plotting area.
3. Change `eval: false` to `true` in the download cell—but replace its URL with
   a real, permitted source first.
4. Add `code-fold: true` under the HTML format.
5. Move `warning: false` from document YAML to one cell and compare its scope.

