---
title: "A reproducible analysis"
subtitle: "Change scores in a synthetic training study"
author: "Workshop participant"
date: today
bibliography: bibliography.bib
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
    html-math-method: mathml
  pdf:
    number-sections: true
execute:
  warning: false
  message: false
---

# Introduction

Literate programming keeps explanation and computation in the same source
document [@knuth1984]. This reduces opportunities for reported values, figures,
and tables to become detached from the analysis that produced them.

## Research question

How did scores change between baseline and follow-up, and how did the descriptive
change differ between the comparison and intervention groups?

The analysis aims to:

- calculate an individual change score;
- summarise change within each study group; and
- visualise the association between training hours and change.

For participant $i$, the change score is:

$$
\Delta_i = y_{i,\mathrm{followup}} - y_{i,\mathrm{baseline}}.
$$

::: {.callout-note}
The data are entirely synthetic. They contain no real participants and cannot
provide evidence about a real intervention.
:::

# Methods

The CSV is read from a relative path, and the change score is derived in code.

```{r}
#| label: load-and-prepare
#| include: false

library(readr)
library(dplyr)
library(ggplot2)

study <- read_csv("data/training-study.csv", show_col_types = FALSE) |>
  mutate(change = followup_score - baseline_score)
```

The dataset contains `r nrow(study)` fictional participants, with
`r sum(study$group == "comparison")` in the comparison group and
`r sum(study$group == "intervention")` in the intervention group.

# Results

```{r}
#| label: tbl-summary
#| tbl-cap: "Change scores by study group."

summary_table <- study |>
  group_by(group) |>
  summarise(
    participants = n(),
    mean_change = mean(change),
    sd_change = sd(change),
    .groups = "drop"
  )

knitr::kable(summary_table, digits = 1)
```

@tbl-summary shows a larger average change in the intervention group than in
the comparison group.

```{r}
#| label: fig-change
#| fig-cap: "Change score by training hours and study group."
#| fig-alt: "Scatterplot of training hours against change score. Comparison observations cluster at fewer hours and smaller changes; intervention observations generally have more hours and larger changes."

ggplot(study, aes(training_hours, change, colour = group, shape = group)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Training hours",
    y = "Change in score",
    colour = "Study group",
    shape = "Study group"
  ) +
  theme_minimal()
```

The constructed relationship in @fig-change shows that group membership,
training hours, and change are all associated in this synthetic dataset.

# Discussion

The intervention group has the larger descriptive mean change. This comparison
does not establish that the intervention caused the difference: the data are
synthetic, participants were not randomised, and training hours differ
systematically between groups. A real analysis would require a design-appropriate
model, uncertainty estimates, and substantive assessment of possible
confounding.

Reproducible computation keeps the numerical summary aligned with the source
data, but it does not remove the author's responsibility for design,
interpretation, or transparent limitations [@sandve2013].

# Reproducibility

The source data, transformations, prose, and output instructions are stored in
one project. The following information records the R environment used for this
render.

```{r}
#| label: session-information
#| code-fold: true
sessionInfo()
```

# References
