# Literate computation {#sec-computation}

An executable document contains three interleaved kinds of material:

1. **narrative**—the question, methods, interpretation, and limitations;
2. **code**—instructions for reading, transforming, analysing, and presenting
   data; and
3. **computed output**—values, messages, tables, and figures created when the
   document renders.

The rendered document is a record of code executed in context, not a screenshot
of an interactive session.

```{r}
#| label: workshop-demo-setup
#| include: false

library(readr)
library(dplyr)
library(ggplot2)

study <- read_csv(
  "materials/complete/data/training-study.csv",
  show_col_types = FALSE
) |>
  mutate(change = followup_score - baseline_score)
```

## Anatomy of an R cell {#sec-chunk-anatomy}

An executable R cell begins with `` ```{r} `` and ends with three backticks.
Cell options sit at the top and begin with `#|`:

```r
#| label: group-counts
#| echo: true
#| message: false

study |>
  count(group)
```

When this book renders, that pattern produces a live demonstration:

```{r}
#| label: demo-group-counts

study |>
  count(group)
```

The cell has:

- an **engine** (`r`);
- a **label** (`demo-group-counts`);
- zero or more **options**;
- R code; and
- output generated during rendering.

Labels should be unique, descriptive, and stable. Labels with reserved prefixes
such as `fig-` and `tbl-` create cross-referenceable outputs.

## The execution model

For an R-based `.qmd`, Quarto asks `knitr` to execute cells from top to bottom.
Objects created in an earlier cell are available later in that render.

```r
# Cell 1
study <- readr::read_csv("data/training-study.csv")

# Cell 2: depends on Cell 1
nrow(study)
```

This has several consequences:

- changing cell order can change whether code works;
- a document should create every object it uses;
- packages should be loaded or namespaced in the source;
- interactive Global Environment objects are not reliable dependencies; and
- restarting R before rendering is a useful test.

### A common hidden-state failure

Imagine running this interactively:

```r
study_with_change <- study |>
  mutate(change = followup_score - baseline_score)
```

You then write a document cell containing only:

```r
mean(study_with_change$change)
```

It works until R is restarted. The document contains the result-producing line
but not the instruction that creates its input. A clean render reveals the
missing step.

## Cell options control execution and presentation {#sec-cell-options}

Chunk options are not cosmetic trivia. They determine what runs, what readers
see, what can be referenced, and how output is presented.

### Core execution and visibility options

| Option | Typical value | Code runs? | Code shown? | Output shown? |
|---|---:|:---:|:---:|:---:|
| `eval` | `false` | No | Yes | No |
| `echo` | `false` | Yes | No | Yes |
| `output` | `false` | Yes | Yes | No |
| `include` | `false` | Yes | No | No |
| `warning` | `false` | Yes | As configured | Warnings hidden |
| `message` | `false` | Yes | As configured | Messages hidden |
| `error` | `true` | Yes | As configured | Error displayed; render continues |

These options answer different questions:

- **Should the code run?** `eval`
- **Should readers see the code?** `echo`
- **Should readers see its normal output?** `output`
- **Should anything from the cell appear?** `include`
- **How should diagnostics be handled?** `warning`, `message`, `error`

### Default: code and output

Use the default when the calculation itself is part of the explanation:

```{r}
#| label: demo-code-and-output

study |>
  summarise(
    participants = n(),
    mean_change = mean(change)
  )
```

The reader can inspect the instructions and the result.

### `echo: false`: show the result, hide the implementation

This is common in a polished reader-facing report:

```{r}
#| label: demo-echo-false
#| echo: false

study |>
  count(group)
```

The code still runs and remains in the `.qmd` source, but the rendered reader
sees only the result. Hiding code is a presentation choice, not reproducibility
by itself; the source must still be available to someone expected to audit it.

### `eval: false`: show an example without running it

```{r}
#| label: demo-eval-false
#| eval: false

# An illustrative call that should not run during this render
download.file("https://example.org/large-data.csv", "data/large-data.csv")
```

This is useful for teaching installation commands, destructive examples, slow
jobs, or code requiring credentials. Do not use it for code that supposedly
created the reported result.

### `output: false`: run code, hide normal output

```{r}
#| label: demo-output-false
#| output: false

summary_table <- study |>
  group_by(group) |>
  summarise(mean_change = mean(change), .groups = "drop")

summary_table
```

The object is created for later cells, and the code is visible, but the printed
table is suppressed here.

### `include: false`: setup that remains out of the document

The first cell in this chapter loaded packages and data using:

```r
#| include: false

library(readr)
library(dplyr)
library(ggplot2)
study <- read_csv("materials/complete/data/training-study.csv")
```

The code ran, but neither code nor output appeared. This is appropriate for
routine setup. It is not a place to conceal analytically important decisions.

### Warnings and messages

Package startup messages rarely help a manuscript reader:

```r
#| message: false
#| warning: false
```

Suppress expected noise only after understanding it. A convergence warning,
failed transformation, or removed observation may be evidence of an analysis
problem. Fix or explain substantive warnings rather than hiding them globally.

### Folding rather than removing code

HTML can keep code available without placing it in the main reading flow:

```yaml
format:
  html:
    code-fold: true
    code-summary: "Show the R code"
```

This affects HTML interaction. PDF cannot fold code, so decide separately
whether code should be shown, hidden, or moved to an appendix.

## Where options can be set

Options can apply at three levels.

### Project level

In `_quarto.yml`:

```yaml
execute:
  warning: false
  message: false
```

Use project defaults for consistent policy across documents.

### Document level

In a document's YAML:

```yaml
---
execute:
  echo: true
  warning: false
---
```

Use document defaults for the intended audience of that document.

### Cell level

```r
#| echo: false
#| fig-width: 7
```

Cell options override broader defaults for one case. The practical precedence is:

```text
project default < document default < cell option
```

Keep exceptions local and intentional. If every cell repeats the same option,
move it to document or project YAML.

## Inline computation

Inline code places a calculated value inside a sentence. The source pattern is:

<code>The study contains &#96;r nrow(study)&#96; participants.</code>

In this rendered chapter, the live sentence is:

> The study contains `r nrow(study)` participants.

Other source examples include:

- The mean change was
  <code>&#96;r round(mean(study$change), 1)&#96;</code> points.
- There were
  <code>&#96;r dplyr::n_distinct(study$group)&#96;</code> study groups.

Use short, readable expressions inline. If a calculation is complicated, create
a clearly named object in a cell and insert only that object:

```r
overall_mean_change <- mean(study$change)
```

The prose can then use:

<code>The mean change was &#96;r round(overall_mean_change, 1)&#96; points.</code>

This keeps prose readable and makes the calculation easier to test.

## Figures, options, and cross-references {#sec-computational-figures}

A figure cell combines execution options with publication metadata:

```r
#| label: fig-change-demo
#| fig-cap: "Change score by training hours and study group."
#| fig-alt: "Scatterplot of training hours against change score..."
#| fig-width: 7
#| fig-height: 4.5
#| dpi: 300

ggplot(study, aes(training_hours, change, colour = group, shape = group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

The live output is @fig-change-demo:

```{r}
#| label: fig-change-demo
#| fig-cap: "Change score by training hours and study group."
#| fig-alt: "Scatterplot of training hours against change score. Comparison observations cluster at fewer hours and smaller changes; intervention observations generally have more hours and larger changes."
#| fig-width: 7
#| fig-height: 4.5
#| dpi: 300

ggplot(study, aes(training_hours, change, colour = group, shape = group)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Training hours",
    y = "Change in score",
    colour = "Study group",
    shape = "Study group"
  ) +
  theme_minimal()
```

Important figure options include:

| Option | Purpose |
|---|---|
| `label: fig-name` | Creates a stable cross-reference target |
| `fig-cap` | Visible caption |
| `fig-alt` | Accessible description of information in the figure |
| `fig-width`, `fig-height` | Size of the plotting device in inches |
| `dpi` | Raster resolution |
| `fig-align` | Alignment in the output |
| `fig-show: hold` | Holds multiple plots from one cell together |

Plot dimensions affect the space in which the plot is drawn, not merely how a
finished image is stretched. Text, facets, and legends may reflow when dimensions
change.

## Tables and cross-references {#sec-computational-tables}

Create the data object in code:

```{r}
#| label: create-summary-table
#| include: false

group_summary <- study |>
  group_by(group) |>
  summarise(
    participants = n(),
    mean_change = mean(change),
    sd_change = sd(change),
    .groups = "drop"
  )
```

Then present it with a caption and table label:

```{r}
#| label: tbl-change-demo
#| tbl-cap: "Change scores by study group."

knitr::kable(group_summary, digits = 1)
```

@tbl-change-demo is numbered and referenced from prose. If another table is
inserted earlier, Quarto updates the number.

Keep a distinction between:

- the **data object** (`group_summary`);
- the **presentation function** (`knitr::kable()`); and
- the **interpretation** in prose.

This makes it easier to change table styling without recalculating the result.

## Results produced “as is”

Some R functions deliberately return Markdown or HTML. With knitr, `results:
'asis'` or Quarto's `output: asis` can ask the renderer to interpret returned
markup rather than print it literally.

Use this only when the code is intentionally generating document structure.
Untrusted text must not be inserted as raw HTML, and format-specific markup may
fail in PDF or Word.

## Errors during rendering

By default, an error stops the render. This is usually correct: a document with a
failed analysis should not silently produce a polished output.

For a teaching document that intentionally demonstrates an error:

```r
#| error: true

mean("not numeric")
```

The error appears in the rendered lesson and the document continues. Do not set
`error: true` globally in a production manuscript merely to obtain an output.

## Randomness

Simulations, sampling, model initialisation, and some algorithms use random
numbers. Without a seed, repeated renders may differ:

```r
set.seed(20260729)
sample(1:100, 5)
```

A seed makes the pseudorandom sequence repeatable. It does not make a simulation
design appropriate or guarantee identical results across every software
implementation.

## Caching and freezing

Two related features can avoid unnecessary computation:

- **cell caching** can reuse a cell's previous computational result; and
- **Quarto freeze** can reuse rendered computational Markdown for a document.

Examples:

```yaml
execute:
  cache: true
```

```yaml
execute:
  freeze: auto
```

These improve rendering speed, but cached output can conceal missing external
dependencies during testing. Before release, perform at least one clean,
uncached render in a fresh copy or controlled environment.

Never use caching as a substitute for recording inputs.

## Choosing options for the reader

The same analysis can be presented differently:

| Audience | Reasonable presentation |
|---|---|
| Analyst collaborator | Show code and output |
| Journal reader | Hide routine setup; show methods-critical code selectively |
| Auditor or reviewer | Provide source or folded code plus environment details |
| Student | Show starter code; reserve answers for solution output |
| Lecture audience | Hide most code and enlarge the explanatory figure |

Chunk options let one computational workflow support these audiences, but a
document should not pretend that hidden source does not exist.

## A substantial computation exercise

In `manuscript.qmd`:

1. Add an `include: false` setup cell that loads packages and reads the CSV.
2. Derive `change` in code.
3. Replace typed sample-size and group-count values with inline R.
4. Create a summary object in a hidden setup cell.
5. Present it in a `tbl-` cell with a caption and refer to it from prose.
6. Create a `fig-` cell with a caption, informative alt text, explicit
   dimensions, and both colour and shape mappings.
7. Use `code-fold` for HTML and decide explicitly how code appears in PDF.
8. Restart R and render without running any cells manually.
9. Change one follow-up value in a temporary copy of the CSV and list every
   output that updates.
10. Restore the source data and render again.

::: {.checkpoint}
### Checkpoint

You can explain the difference between `eval`, `echo`, `output`, and `include`;
identify which cell creates every object; and trace a reported value back to the
source data and code that generated it.
:::
